---
title: "DATA 3230 Data Visualization Final Project"
format:
  pdf:
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

# Introduction

This project analyzes monthly MTA major incidents in New York City and examines whether weather conditions help explain variation in incident counts. Major incidents are defined as incidents that delay 50 or more trains. The analysis begins with weather-only models, then improves the model by incorporating subway ridership as a measure of system usage, and finally evaluates whether logarithmic transformations provide a better representation of the relationship between incidents, weather, and ridership.

# Load Packages

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(httr)
library(car)
```

# Load and Summarize Incident Data

```{r}
incidents_2015_2019 <- read_csv("~/Downloads/MTAIncidents 2015-2019.csv", show_col_types = FALSE)
incidents_2020_2024 <- read_csv("~/Downloads/MTAIncidents 2020-2024.csv", show_col_types = FALSE)

all_incidents <- bind_rows(incidents_2015_2019, incidents_2020_2024) %>%
  arrange(month)

summary_df <- all_incidents %>%
  group_by(month, category) %>%
  summarise(count = n(), .groups = "drop")

monthly_totals <- all_incidents %>%
  group_by(month) %>%
  summarise(total = n(), .groups = "drop") %>%
  arrange(month)

category_totals <- all_incidents %>%
  group_by(category) %>%
  summarise(total = n(), .groups = "drop") %>%
  arrange(desc(total))
```

# Download and Prepare Weather Data

```{r}
base_url <- "https://www.ncei.noaa.gov/access/services/data/v1"

res <- GET(
  url = base_url,
  query = list(
    dataset = "daily-summaries",
    stations = "USW00094728",
    startDate = "2015-01-01",
    endDate = "2024-12-31",
    dataTypes = "TMAX,TMIN,PRCP,SNOW,SNWD",
    format = "json",
    units = "standard",
    includeAttributes = "false"
  ),
  timeout(60)
)

stop_for_status(res)

weather_daily <- content(res, as = "parsed", type = "application/json") %>%
  bind_rows() %>%
  mutate(
    DATE = as.Date(DATE),
    TMAX = as.numeric(TMAX),
    TMIN = as.numeric(TMIN),
    PRCP = as.numeric(PRCP),
    SNOW = as.numeric(SNOW),
    SNWD = as.numeric(SNWD),
    month = floor_date(DATE, "month"),
    heavy_precip_day = PRCP >= 0.5,
    hot_day = TMAX >= 90
  )

weather_monthly <- weather_daily %>%
  group_by(month) %>%
  summarise(
    avg_tmax = mean(TMAX, na.rm = TRUE),
    avg_tmin = mean(TMIN, na.rm = TRUE),
    total_prcp = sum(PRCP, na.rm = TRUE),
    total_snow = sum(SNOW, na.rm = TRUE),
    heavy_precip_days = sum(heavy_precip_day, na.rm = TRUE),
    hot_days = sum(hot_day, na.rm = TRUE),
    n_days = n(),
    .groups = "drop"
  ) %>%
  arrange(month)
```

# Combine Incidents and Weather Data

```{r}
combined <- monthly_totals %>%
  mutate(month = as.Date(month)) %>%
  left_join(weather_monthly, by = "month")
```

# Correlation Analysis

```{r}
numeric_vars <- combined %>%
  select(
    total, avg_tmax, avg_tmin, total_prcp, total_snow,
    heavy_precip_days, hot_days, n_days
  )

cor_mat <- cor(numeric_vars, use = "pairwise.complete.obs")
cor_mat
```

The correlation matrix shows that snowfall has the strongest relationship with incidents among the weather variables, while most other weather measures have weak associations.

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

cor_long <- as.data.frame(cor_mat) %>%
  mutate(var1 = rownames(.)) %>%
  pivot_longer(
    cols = -var1,
    names_to = "var2",
    values_to = "correlation"
  )

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = NULL,
    y = NULL,
    fill = "Correlation",
    title = "Correlation Heatmap: Monthly MTA Incidents vs. NYC Weather"
  )
```

The heatmap confirms that snowfall is the weather feature most visibly associated with incident totals.

# Initial Regression Models Using Weather Only

```{r}
model1 <- lm(total ~ avg_tmax + total_prcp + total_snow, data = combined)
summary(model1)

model2 <- lm(total ~ hot_days + heavy_precip_days + total_snow, data = combined)
summary(model2)
```

These weather-only models show that snowfall is significant, but the overall explanatory power is low.

```{r}
vars_for_cor <- combined %>%
  select(avg_tmax, avg_tmin, total_prcp, total_snow, heavy_precip_days, hot_days)

cor(vars_for_cor, use = "pairwise.complete.obs")

vif(model1)
vif(model2)
```

The low VIF values indicate that multicollinearity is not a major concern in these models.

# Simple Linear Regression with Snow Only

```{r}
snow_model <- lm(total ~ total_snow, data = combined)
summary(snow_model)
```

The snow-only model shows that snowfall is a significant predictor, but it explains only a small portion of the variation in incidents.

## Q-Q Plot for Snow-Only Model

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

qqnorm(residuals(snow_model))
qqline(residuals(snow_model), col = "red", lwd = 2)
```

The residuals are approximately normal, with some deviation in the tails.

## Residuals vs Fitted for Snow-Only Model

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(
  fitted(snow_model),
  residuals(snow_model),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)
abline(h = 0, col = "red", lwd = 2)
```

The residual pattern suggests the snow-only model misses important variation in the data.

# Log-Transformed Snow-Only Model

```{r}
log_snow_model <- lm(log(total) ~ total_snow, data = combined)
summary(log_snow_model)

```

The log transformation changes the relationship slightly, but the model still leaves substantial variation unexplained because it does not yet account for differences in subway usage.

# Add Subway Ridership Data (2020 and Later)

```{r}
ridership <- read_csv("~/Downloads/MTA RIDERSHIP SINCE 2020.csv", show_col_types = FALSE)

ridership_monthly <- ridership %>%
  filter(Mode == "Subway") %>%
  mutate(
    Date = mdy(Date),
    month = floor_date(Date, "month")
  ) %>%
  group_by(month) %>%
  summarise(
    ridership = sum(Count, na.rm = TRUE),
    .groups = "drop"
  )
```

Ridership is aggregated to the monthly level so it matches the monthly incident data.

```{r}
combined_2020 <- combined %>%
  mutate(month = as.Date(month)) %>%
  filter(month >= as.Date("2020-01-01")) %>%
  left_join(ridership_monthly, by = "month") %>%
  filter(!is.na(ridership)) %>%
  mutate(ridership_millions = ridership / 1e6)
```

Restricting to 2020 onward allows incidents, weather, and ridership to be analyzed on the same time scale.

# Linear Model with Snow and Ridership

```{r}
final_model <- lm(total ~ total_snow + ridership_millions, data = combined_2020)
summary(final_model)
```

The linear model with snow and ridership substantially improves on the earlier weather-only models. Both snowfall and ridership are statistically significant predictors of monthly major incidents.

An increase of 1 million riders is associated with additional major incidents, holding snowfall constant. Likewise, each additional inch of snow is associated with more incidents, holding ridership constant.

# Incident Rate Model

```{r}
combined_2020 <- combined_2020 %>%
  mutate(incident_rate = total / ridership)

rate_model <- lm(incident_rate ~ total_snow, data = combined_2020)
summary(rate_model)
```

The incident rate model shows that snowfall remains significant even after adjusting for ridership.

# Time Series of Incidents and Ridership

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

ggplot(combined_2020, aes(x = month)) +
  geom_line(aes(y = total, color = "Incidents"), linewidth = 1) +
  geom_line(aes(y = ridership_millions * 5, color = "Ridership (scaled)"), linewidth = 1) +
  scale_color_manual(values = c("Incidents" = "black", "Ridership (scaled)" = "blue")) +
  theme_minimal() +
  labs(
    title = "Monthly Major Incidents and Subway Ridership Over Time",
    x = "Month",
    y = "Incidents (Ridership scaled for comparison)",
    color = ""
  )
```

Incidents and ridership move together over time, suggesting that higher system usage is associated with more disruptions.

# Diagnostics for Linear Model

## Residuals vs Fitted

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(final_model, which = 1)
```

## Normal Q-Q

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(final_model, which = 2)
```

## Scale-Location

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(final_model, which = 3)
```

## Residuals vs Leverage

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(final_model, which = 5)
```

```{r}
vif(final_model)

```

The linear model performs better than the earlier models, but there is still some evidence that the relationship may not be perfectly linear.

# Model Comparison with Logarithmic Transformations

To further improve the model, logarithmic transformations were considered. A log-transformed response can reduce skewness and stabilize variance, while a log-log model can capture multiplicative relationships and allow coefficients to be interpreted as elasticities.

```{r}
model_linear <- lm(total ~ total_snow + ridership_millions, data = combined_2020)

model_log <- lm(log(total) ~ total_snow + ridership_millions, data = combined_2020)

model_loglog <- lm(log(total) ~ log(total_snow + 1) + log(ridership_millions), data = combined_2020)

AIC(model_linear, model_log, model_loglog)
summary(model_loglog)
```

The AIC comparison shows that the log-log model provides the best fit among the three specifications. This suggests that the relationship between incidents, snowfall, and ridership is better represented on a multiplicative scale than on a purely linear one.

# Additional Visualizations for the Log-Log Model

## Log-Log Relationship: Snowfall vs Incidents

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

ggplot(combined_2020, aes(x = log(total_snow + 1), y = log(total))) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(
    title = "Log-Log Relationship: Snowfall vs. Monthly Major Incidents",
    x = "log(Total Snow + 1)",
    y = "log(Monthly Major Incidents)"
  )
```

This plot shows the transformed relationship between snowfall and incidents. On the log scale, the pattern appears more linear than in the untransformed model.

## Log-Log Relationship: Ridership vs Incidents

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

ggplot(combined_2020, aes(x = log(ridership_millions), y = log(total))) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(
    title = "Log-Log Relationship: Ridership vs. Monthly Major Incidents",
    x = "log(Ridership in Millions)",
    y = "log(Monthly Major Incidents)"
  )
```

This plot shows that ridership also has a clearer linear relationship with incidents after the logarithmic transformation.

## Actual vs Predicted Monthly Incidents

```{r}
combined_2020 <- combined_2020 %>%
  mutate(
    predicted_loglog = predict(model_loglog),
    predicted_total = exp(predicted_loglog)
  )

```

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

ggplot(combined_2020, aes(x = total, y = predicted_total)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Actual vs Predicted Monthly Incidents",
    x = "Actual Incidents",
    y = "Predicted Incidents"
  )
```

If the model fit perfectly, the points would lie on the dashed line. The plot shows that the log-log model tracks the observed data reasonably well.

## Actual vs Predicted Over Time

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

ggplot(combined_2020, aes(x = month)) +
  geom_line(aes(y = total, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = predicted_total, color = "Predicted"), linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Actual vs Predicted Monthly Major Incidents Over Time",
    x = "Month",
    y = "Monthly Major Incidents",
    color = ""
  )
```

This time series comparison shows how closely the fitted values from the log-log model follow the observed monthly incident totals.

# Diagnostics for the Log-Log Model

## Residuals vs Fitted

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(model_loglog, which = 1)
```

## Normal Q-Q

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(model_loglog, which = 2)
```

## Scale-Location

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(model_loglog, which = 3)
```

## Residuals vs Leverage

```{r}
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

plot(model_loglog, which = 5)
```

These diagnostics can be compared to the earlier linear-model diagnostics to assess whether the logarithmic transformation improved residual behavior and overall fit.

# Conclusion

The analysis began with weather-only models, which showed that snowfall was the most meaningful weather predictor, while temperature and precipitation variables were generally not statistically significant. However, these early models had relatively low explanatory power.

A snow-only model confirmed that snowfall mattered, but it still left much of the variation in incidents unexplained. A log-transformed version of that model slightly improved the structure of the relationship, suggesting that transformation could be useful.

After incorporating subway ridership and restricting the analysis to 2020 onward, model performance improved substantially. The linear model with snowfall and ridership showed that both predictors were statistically significant, indicating that both operational exposure and weather conditions contribute to MTA disruptions.

Finally, comparing the linear, log, and log-log models showed that the log-log specification provided the best overall fit according to AIC. The additional visualizations and diagnostics suggest that the transformed model better captures the relationship between incident counts, snowfall, and ridership. Overall, the results suggest that monthly MTA major incidents are better explained by a multiplicative relationship in which both system usage and weather intensity play important roles.
